Evaluating the Decency and Consistency of Data Validation Tests Generated by LLMs

### 2023-10-01 paper/paper.qmd
lindsay edits

Much important work has been done to illustrate the importance of reproducibility and reliability in data science across domains. In particular reproducibility promotes transparency, trustworthiness and credibility, and a more thorough understanding of the data. The value of data validation is not limited to particular domains---it is a fundamental component of a reliable data science workflow, regardless of the data at hand.

There are a number of challenges that accompany the development and implementation of data validation tests. First, implementing automated tests requires the ability to develop code [@zhang20]. While there are a number of accessible, well-documented testing-centered libraries in different programming languages (e.g., `Great Expectations` in Python, and `pointblank` in R), the implementation of a test suite may require significant data processing work. One of the main takeaways from @katzandmoore on their data validation efforts for the IJF is that data tests should not be compromised just to be compatible with pre-defined test functions that exist in a software package. In order to develop the most valuable suite of tests, it may be necessary to create new variables in a dataset or to merge multiple datasets, for example. This presents a challenge for individuals who are not trained in or comfortable with computer programming, but wish to programmatically test their data. In addition, the development of a comprehensive, bespoke, and accurate data test suite can rarely be done by one data scientist alone. As advocated by @katzandmoore, domain expertise is fundamental for the creation and development of high-quality data tests. For instance, without incorporating the knowledge that donation date is only collected in three jurisdictions (British Columbia, Ontario and Federal) into their code, @katzandmoore's test surrounding missingness for that variable would have produced inflated failure rates. This is knowledge that was gained through collaborative efforts with IJF team members who were involved in dataset construction. For an individual working on an unfamiliar dataset, seeking out domain knowledge and gaining a thorough understanding of one's data is a non-trivial task in and of itself. Finally, the process of designing, developing, and deploying data validation tests is iterative in nature and takes a great deal of time. For researchers or journalists who are eager to publish their findings for instance, there may not be enough incentive or resources for them to want to do this time-consuming validation work. This challenge is only amplified if an individual was not involved in the dataset construction process and does not have access to vital domain expertise, or if an individual does not have programming experience.

Evidently, while data validation is incredibly valuable for the production of transparent and trustworthy research, it comes with many challenges which hinder its accessibility. This work presents an opportunity to promote data validation through the use of LLMs. While there has been a great deal of work done to develop more accessible programming tools for data validation, these tools do not entirely mitigate the challenges which accompany data validation, namely the need for programming knowledge, domain expertise, and a great deal of time. Our findings illustrate the ability of LLMs to rapidly produce usable data validation tests, written in code, with limited information.

@@

Our work is consistent with previous literature demonstrating that LLM performance on complex, user-defined tasks is sensitive to prompt engineering. Prompting is a brittle process in which minor alterations can lead to large fluctuations in performance [@arora2022, @zhao2021], however with OpenAI’s LLMs in particular, GPT-4 [@giorgi2023] seems less susceptible to this than ChatGPT (GPT-3.5). In general, previous research has demonstrated that when using ChatGPT, prompts should provide context [@white2023, @clavie2023], be specific [@white2023, @yong2023, @openaiprompt], define a persona or role [@white2023, @clavie2023], break complex reasoning into smaller steps [@white2023, @henrickson2023], and provide example responses [@brown2020, @arora2022, @openaiprompt] to improve performance on a variety of tasks, including classifying job types [@clavie2023], generating images for construction defect detection [@yong2023], and generating hermeneutically valuable text [@henrickson2023].

These observations are predominantly borne out in our experiments. For instance, the more specified prompt (“Describe”) and the prompt eliciting step-by-step thinking (“Simulate”) both show improvements over the simplest prompt (“Name”). However, improvements made by specificity break down when example data is provided (“Example”). We hypothesize that this is due to ChatGPT and GPT-4’s poor performance on numerical and structured data in comparison to more domain-specific LLMs such as BloombergGPT which has been trained on numeric, structured financial data and utilizes a tokenizer more suited to these data [@wu2023]. We do not see significant changes in performance when two different roles are specified to the LLM, which contradicts previous literature [@white2023, @clavie2023]. Finally, our significant improvements moving from zero- to one-shot and marginal to no improvements from one- to few-shot align with existing literature [@giorgi2023].  Some studies have demonstrated ChatGPT and GPT-4’s difficulties in responding to queries in the desired format during zero-shot structured tasks, including named entity recognition [@hu2023] and multiple choice question [@clavie2023]. @giorgi2023 thus attribute performance improvements in one- and few-shot prompts to the in-context examples’ guidance as to the desired output structure. Upon inspection of the outputs from zero-, one-, and few-shot prompts, we believe that this same effect exists for our prompts.

Our work additionally contributes to the growing body of work using LLMs to automate portions of the data science workflow. LLMs have already been integrated into software tools such as GitHub's Co-Pilot [@githubcopilot] and AutoGPT [@autogpt] and IDEs such as IntelliJ [@krochmalski2014] and VSCode [@vscode] to produce code in a variety of languages. These tools are and will continue to assist data scientists to accelerate, resolve bugs in, document, and optimize their code in all parts of the data science workflow. LLMs have also demonstrated potential to automate other onerous or tedious tasks in data science pipelines. These include generating synthetic text data [@chung2023, @yu2023], improving search-based software testing [@lemieux2023], generating unit tests from natural language in a variety of programming languages and contexts [@lahiri2022, @schafer2023], generating property-based tests from API documentation [@vikram2023large], and conducting exploratory data analysis [@ma2023]. As LLM prompting and fine-tuning methods improve, they will reduce technical barriers to conducting transparent, accurate, reproducible analysis, enabling data scientists to be more productive and put their energy towards analysis. higher-order reasoning, and sophisticated inference. The work described in this paper towards automating data validation using LLMs has the capacity to further this evolution.

>>>>

Data validation is a cornerstone of high quality data science workflows. Broadly, data validation is centered around establishing the credibility of individual data values, the coherence of data with itself, and the consistency of data with relevant external sources [@alexander2023]. Data quality — or a lack thereof — can have fundamental downstream effects on the rest of the scientific process. If the data being supplied to a statistical model is incorrect in some way (e.g., a categorical variable is stored as numeric), this will have significant implications for subsequent results and conclusions [@alexander2023; @hynes17; @breck2019; @zhang20; @gao2016]. A powerful tool to implement data validation in practice is automated testing, which intuits data expectations from format, column name, or other meta information and converts those expectations to code [@alexander2023; @zhang20]. Automated validation promotes trustworthiness and transparency. By developing a suite of tests to validate that particular characteristics of a dataset hold true, data scientists uncover fundamental assumptions that underpin their work and equip users with confidence that those assumptions are met in practice. The value of data validation testing is not limited to particular domains --- it is a fundamental component of a reliable data science workflow, regardless of the data at hand [@alexander2023; @gao2016].

There are a number of challenges that accompany the development and implementation of data validation tests. These challenges are particularly important to consider when thinking about data validation efforts across research domains and levels of data science expertise.

First, implementing automated tests requires the ability to develop code. There are a number of testing-centered libraries in various programming languages. Great Expectations — the Python library employed by @katzandmoore — contains various data testing functions and has a data assistant functionality which programmatically explores data to develop tests based on observed characteristics [@gx23]. `pointblank` is an R package which offers predefined automated data testing functionality [@pointblank] and IBM SPSS statistics can perform data validation checks by identifying unexpected or invalid entries in a dataset [@IBM23]. With a focus on data validation for machine learning (ML), @hynes17 developed a data linter tool that detects possible inconsistencies in the data and suggests appropriate transformations for a specific model type. Further, @breck2019 executed an automated validation ML platform at Google to address concerns about the downstream effects of messy data in model training. While various data testing software have been developed and shared, many of these have limited functionality, and the implementation of a test suite may require significant data processing work and necessarily requires programming knowledge [@alexander2023; @zhang20]. From their work on data validation for the IJF @katzandmoore conclude that data tests should not be compromised just to be compatible with predefined test functions that exist in a software package. In order to develop the most valuable suite of tests, it may be necessary to create new variables in a dataset, merge multiple datasets, or create tests which are outside the scope of existing software frameworks, for example. This presents a challenge for individuals who are not trained in or comfortable with computer programming, but wish to programmatically test their data.

In addition, the development of a comprehensive, bespoke, and accurate data test suite can rarely be done by one data scientist alone. As advocated by @katzandmoore, domain expertise is fundamental to the creation and development of high-quality data tests. For instance, without incorporating the knowledge that donation dates are only collected in three jurisdictions (British Columbia, Ontario and Federal) into their code, @katzandmoore's test for missingness for that variable would have produced inflated failure rates. This is knowledge that was gained through collaborative efforts with IJF team members who were involved in dataset construction. For an individual working on an unfamiliar dataset, seeking out domain knowledge and gaining a thorough understanding of that data is a non-trivial task in and of itself.

Finally, the process of designing, developing, and deploying data validation tests is iterative in nature and takes a great deal of time [@katzandmoore]. For researchers or journalists who are eager to publish their findings, for instance, there may not be enough incentive or resources for them to do this time-consuming validation work. This challenge is only amplified if an individual was not involved in the dataset construction process, does not have access to vital domain expertise, or does not have programming experience.

While data validation is incredibly valuable for the production of transparent and trustworthy research, it comes with many challenges which hinder its accessibility. This work presents an opportunity to advocate for an alternative approach to data validation through the use of LLMs. Though there has been a great deal of work done to develop more accessible programming tools for data validation, these tools do not entirely mitigate the challenges which accompany the process, namely the need for programming knowledge, domain expertise, and a great deal of time. Our findings illustrate the potential for LLMs to rapidly produce usable data validation tests, written in code, with limited information --- a promising application to support the promotion of more trustworthy, transparent data workflows across domains.

@@

Our work is consistent with previous literature demonstrating that LLM performance on complex, user-defined tasks is sensitive to prompt engineering. Prompting is a brittle process in which minor alterations can lead to large fluctuations in performance [@arora2022; @zhao2021], however with OpenAI’s LLMs in particular, GPT-4 [@giorgi2023] seems less susceptible to this than ChatGPT (GPT-3.5). In general, previous research has demonstrated that when using ChatGPT, prompts should provide context [@white2023; @clavie2023], be specific [@white2023; @yong2023; @openaiprompt], define a persona or role [@white2023; @clavie2023], break complex reasoning into smaller steps [@white2023; @henrickson2023], and provide example responses [@brown2020; @arora2022; @openaiprompt] to improve performance on a variety of tasks, including classifying job types [@clavie2023], generating images for construction defect detection [@yong2023], and generating hermeneutically valuable text [@henrickson2023].

These observations are predominantly borne out in our experiments. For instance, the more specified prompt (“Describe”) and the prompt eliciting step-by-step thinking (“Simulate”) both show improvements over the simplest prompt (“Name”). However, improvements made by specificity break down when example data is provided (“Example”). We hypothesize that this is due to ChatGPT and GPT-4’s poor performance on numerical and structured data in comparison to more domain-specific LLMs such as BloombergGPT which has been trained on numeric, structured financial data and utilizes a tokenizer more suited to these data [@wu2023]. We do not see significant changes in performance when two different roles are specified to the LLM, which contradicts previous literature [@white2023; @clavie2023]. Finally, our significant improvements moving from zero- to one-shot and marginal to no improvements from one- to few-shot align with existing literature [@giorgi2023].  Some studies have demonstrated ChatGPT and GPT-4’s difficulties in responding to queries in the desired format during zero-shot structured tasks, including named entity recognition [@hu2023] and multiple choice question [@clavie2023]. @giorgi2023 thus attribute performance improvements in one- and few-shot prompts to the in-context examples’ guidance as to the desired output structure. Upon inspection of the outputs from zero-, one-, and few-shot prompts, we believe that this same effect exists for our prompts.

Our work additionally contributes to the growing body of work using LLMs to automate portions of the data science workflow. LLMs have already been integrated into software tools such as GitHub's Co-Pilot [@githubcopilot] and AutoGPT [@autogpt] and IDEs such as IntelliJ [@krochmalski2014] and VSCode [@vscode] to produce code in a variety of languages. These tools are and will continue to assist data scientists to accelerate, resolve bugs in, document, and optimize their code in all parts of the data science workflow. LLMs have also demonstrated potential to automate other onerous or tedious tasks in data science pipelines. These include generating synthetic text data [@chung2023; @yu2023], improving search-based software testing [@lemieux2023], generating unit tests from natural language in a variety of programming languages and contexts [@lahiri2022, @schafer2023], generating property-based tests from API documentation [@vikram2023large], and conducting exploratory data analysis [@ma2023]. As LLM prompting and fine-tuning methods improve, they will reduce technical barriers to conducting transparent, accurate, reproducible analysis, enabling data scientists to be more productive and put their energy towards analysis. higher-order reasoning, and sophisticated inference. The work described in this paper towards automating data validation using LLMs has the capacity to further this evolution.
